---
title: "Electricity Market Prices - July Report"
author: "Egecan Esen"
date: "8/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(ggplot2)
library(lubridate)
raw_df <- read_excel("ptf-smf.xls")
```

## First Glimpse At the Data
This is a report about the electricity prices in July 2020.
The data used in this project can be reached from this [link](https://rapor.epias.com.tr/rapor/xhtml/ptfSmfListeleme.xhtml).

First, let's take a quick look at the data: 

```{r glimpse, echo=TRUE}
raw_df %>% glimpse()
```

## Rearranging the Dataframe
It can be seen that all the data is in form of 'character'. The name of the columns must be changed, 'Tarih' column needs to be converted to 'POSIXct', 'PTF' & 'SMF' need to be converted to 'double'.

```{r, echo=TRUE}
df <- raw_df
df$DateTime <- dmy_hm(df$Tarih)

df[,c(2:4)] <- lapply(df[,c(2:4)],function(x) as.numeric(gsub(",", ".", gsub("\\.", "", x))))

colnames(df)[2] = "MCP" #Market Clearing Price
colnames(df)[3] = "SMP" #System Marginal Price

df <- df %>% select(7,2,3)

df %>% glimpse()
```

The Dataframe is ready to analyze.

## Data Analysis

The visualization of the dataframe: 

```{r, echo=TRUE}
df %>% pivot_longer(.,-DateTime) %>% ggplot(aes(x=DateTime,y=value,color=name)) +geom_line() + theme_minimal() + labs(x="Date", y="TL/MWh", color="", title = "MCP & SMP", subtitle = "Prices during the month of July") + scale_x_datetime(date_breaks = "2 days",date_labels = "%m/%d")
```
```{r, echo=TRUE}
df %>% ggplot(aes(x=MCP,y=SMP, color=DateTime)) +geom_point(position="jitter") + theme_minimal() + labs(title = "MCP vs SMP", subtitle = "Prices during the month of July") + scale_colour_gradientn(name= "Date", colours=rainbow(7), labels= c("30-06","05-07","11-07","17-07","23-07","29-07")) 
```


Let's check some basic daily stats of market clearing and system marginal prices:

```{r echo=TRUE, message=FALSE, warning=FALSE}
#Mean, Median, Maximum, Minimum & Standard Deviation of Each Day's MCP
MCP <- df %>%
  mutate(Day=date(DateTime)) %>%
  group_by(Day) %>%
  summarise(meanMCP = mean(MCP), medianMCP = median(MCP), maxMCP = max(MCP), minMCP = min(MCP), sdMCP =sd(MCP))

#Mean, Median, Maximum, Minimum & Standard Deviation of Each Day's SMP
SMP <- df %>%
  mutate(Day=date(DateTime)) %>%
  group_by(Day) %>%
  summarise(meanSMP = mean(SMP), medianSMP = median(SMP), maxSMP = max(SMP), minSMP = min(SMP), sdSMP = sd(SMP))

Means <- bind_cols(Day = MCP$Day,meanMCP = MCP$meanMCP, meanSMP = SMP$meanSMP)
```

```{r, echo=TRUE}
Means %>% pivot_longer(.,-Day) %>% ggplot(aes(x=Day,y=value,color=name)) +geom_line() + theme_minimal() + labs(x="Date", y="TL/MWh", color="", title = "Daily Average MCP & SMP ", subtitle = "Prices during the month of July") + expand_limits(y=0) + scale_x_date(date_breaks = "1 day", date_labels = "%d")
```

The statistics of hourly prices can be checked in the same way:

```{r echo=TRUE, message=FALSE, warning=FALSE}
df %>%
  mutate(Hour=hour(DateTime)) %>%
  group_by(Hour) %>%
  summarise(meanMCP = mean(MCP), medianMCP = median(MCP), maxMCP = max(MCP), minMCP = min(MCP), sdMCP =sd(MCP))

df %>%
  mutate(Hour=hour(DateTime)) %>%
  group_by(Hour) %>%
  summarise(meanSMP = mean(SMP), medianSMP = median(SMP), maxSMP = max(SMP), minSMP = min(SMP), sdSMP = sd(SMP))
```

From the tables above, it seems that the prices are lower during night time. Below is the comparison of the prices during day (8-22) and night (22-8):

```{r echo=TRUE, message=FALSE, warning=FALSE}
#Daytime hours 
daytime <- df %>%
  mutate(Hour=hour(DateTime),Day=as_date(DateTime)) %>%
  filter(Hour >= 8 & Hour <= 22) %>%
  select(Day,MCP, SMP) %>%
  group_by(Day) %>%
  summarise(MCP = mean(MCP), SMP = mean(SMP), DIFF = round(MCP-SMP,digits=2)) %>%
  arrange(DIFF)

#Nighttime hours
nighttime <- df %>%
  mutate(Hour=hour(DateTime),Night=as_date(DateTime)) %>%
  filter(Hour > 22 | Hour < 8) %>%
  select(Night,MCP, SMP) %>%
  group_by(Night) %>%
  summarise(MCP = mean(MCP), SMP = mean(SMP), DIFF = round(MCP-SMP,digits=2)) %>%
  arrange(DIFF)

daynight <- bind_cols(Date= daytime$Day, Day = daytime$DIFF, Night = nighttime$DIFF)

daynight %>% pivot_longer(.,-Date) %>% ggplot(aes(x=Date,y=value,fill=name)) +geom_bar(stat="identity", position ="dodge") + theme_minimal() + labs(x="Date", y="TL/MWh", color="", title = "Daytime vs Nighttime Difference ", subtitle = "A comparison of average daytime and nighttime MCP&SMP differences ") + scale_x_date(date_breaks = "1 day", date_labels = "%d")

```

Daily Average MCP during daytime is `r format(mean(daytime$MCP))` and standard deviation is `r format(sd(daytime$MCP))` .

Daily Average MCP during nighttime is `r format(mean(nighttime$MCP))` and standard deviation is `r format(sd(nighttime$MCP))` .

Daily Average SMP during daytime is `r format(mean(daytime$SMP))` and standard deviation is `r format(sd(daytime$SMP))` .

Daily Average SMP during nighttime is `r format(mean(nighttime$SMP))` and standard deviation is `r format(sd(nighttime$SMP))` .

It can be seen that the average daytime MCP is higher than SMP,but the reverse is true for SMP.
Also, the standard deviation of SMP is significantly higher than MCP.

The ratio of MCP/SMP is also worth taking a look at:

```{r, echo=TRUE,message=FALSE, warning=FALSE}
#Lowest 10 MCP/SMP ratios
lowest <- df %>%
  mutate(MCP_SMP_RATIO = MCP/SMP) %>%
  select(DateTime,MCP_SMP_RATIO) %>%
  arrange(MCP_SMP_RATIO) %>%
  slice(1:10)

#Highest 10 MCP/SMP ratios
highest <- df %>%
  mutate(MCP_SMP_RATIO = MCP/SMP) %>%
  select(DateTime,MCP_SMP_RATIO) %>%
  arrange(desc(MCP_SMP_RATIO)) %>%
  slice(1:10)

df %>%
  mutate(MCP_SMP_RATIO = MCP/SMP) %>%
  select(DateTime,MCP_SMP_RATIO) %>%
  ggplot(aes(x=DateTime,y=MCP_SMP_RATIO)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 3)) +
  scale_x_datetime(date_breaks = "1 day",date_labels = "%d") +
  labs(x="Date", y="MCP/SMP", color="", title = "MCP/SMP Ratio ") +
  theme_minimal() 
```

The 3 highest MCP/SMP Ratios (`r format(highest$MCP_SMP_RATIO[1:3])`) are significantly higher than the following (`r format(highest$MCP_SMP_RATIO[4:7])` ...) ones.
They are observed consecutively at `r format(highest$DateTime[0:3])` 
.
The same cannot be said about lowest MCP Ratios (`r format(lowest$MCP_SMP_RATIO[1:6])` ...)
The average MCP/SMP Ratio during the month of July is `r format(mean(highest$MCP_SMP_RATIO))`
However, this average drops to `r format(mean(highest$MCP_SMP_RATIO[ -c(1,2,3) ]))` when the 3 highest ratios are removed.

Let's take a look at which & how many hours the MCP is almost equal to SMP:

```{r, echo=TRUE,message=FALSE, warning=FALSE}
almost <- df %>%
  mutate(roundMCP = round(MCP), roundSMP = round(SMP)) %>%
  filter(roundMCP == roundSMP) %>%
  select(DateTime, roundMCP,roundSMP) %>%
  rename(MCP = roundMCP, SMP = roundSMP)
print(almost)
```

The percentage of hours MCP is almost equal to SMP (ratio is almost 1) is `r format(round(nrow(almost)/nrow(df)*100, digits=2))` %

When MCP is higher than SMP, it means there is an energy surplus, when it is lower than SMP, it means there is an energy deficit.

```{r, echo=TRUE,message=FALSE, warning=FALSE}
p <- df %>%
  filter(SMP < MCP)%>%
  count()/count(df)*100
```

```{r, echo=TRUE, eval=TRUE,message=FALSE, warning=FALSE}
df %>% mutate(SD = ifelse(SMP > MCP, "surplus", "deficit")) %>%
  ggplot(aes(x="",y=SD,fill=SD)) + 
  geom_bar(stat="identity",width=1) + 
  coord_polar("y", start = 0) +
  labs(title="Energy deficit/surplus percentage", x="",y="",fill = "Deficit or Surplus") +
  theme_minimal() 
  
```

The percentage of hours where there was an energy deficit: `r format(round(p,digits =3))`
The percentage of hours where there was an energy surplus: `r format(round(100-p,digits=3))`


My progress journal can be reached from this [link](https://pjournal.github.io/boun01-egc-boun/).
